The Ontological Modal Collapse \\ as a Collapse of the Square of Opposition

### 2016-03-25 Papers/2014/Universalis/ModalCollapse.tex
Make minor language improvements

Attempts to prove the
existence (or non-existence) of God by means of abstract, ontological
arguments are an old tradition in western philosophy, with contributions by several prominent philosophers, including St. Anselm of
Canterbury, Descartes and Leibniz. Kurt G{\"o}del and Dana Scott studied and further improved this argument, bringing it to a mathematically more precise form, as a chain of axioms, lemmas and theorems in a second-order modal logic \cite{GoedelNotes,ScottNotes}, shown in Fig. \ref{fig:scott}.

@@

G\"{o}del defines God as a being who possesses all \emph{positive}
properties and states a few reasonable (but debatable) axioms that
such properties should satisfy.  The overall idea of G{\"o}del's proof
is in the tradition of Anselm's argument, who defined God as some
entity of which nothing greater can be conceived.  Anselm argued that
existence in the actual world would make such an assumed being even
greater (more perfect), hence, by definition, God must exist. However,
for Anselm existence was treated as a predicate and the possibility of
God's existence was assumed as granted.  These issues were criticized
by Kant and Leibniz, respectively, and they were addressed in the work
of G\"odel.

Nevertheless, G{\"o}del's work still leaves room for criticism.  In
particular, his axioms are so strong that, when assuming unrestricted
comprehension principles\footnote{A possible direction to remedy modal
  collapse, as studied e.g. by Koons
  \cite{koons06:_sobel_goedel_ontol_proof} is to impose restrictions
  on the domain of properties.}, 
%\marginpar{added footnote}
they entail a \emph{modal collapse}
\cite{Sobel1987,sobel2004logic}: everything that is the case is so
necessarily.  There has been an impressive body of recent and ongoing
work
(cf.~\cite{sobel2004logic,Fitting,anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings,bjordal99,fuhrmann05:_exist_notwen,Hajek2002,Hajek2008,ContemporaryBibliography}
and the references therein) proposing solutions for the modal
collapse.  The goal of this contribution is to discuss the modal
collapse from the point of view of the modal square of opposition. 
Ontological arguments typically rely on an inversion of the normal direction of entailment
in the modal square of opposition for one particular proposition (i.e. God's existence), 
and the modal collapse shows that this inversion in fact occurs for all propositions, resulting in a total collapse of the modal square of opposition.

@@

The modal collapse is an interesting example of philosophical 
controversy and dispute, to which we can apply Leibniz's idea 
of a \emph{calculus ratiocinator} brought to reality in the 
form of contemporary automated theorem provers. 
A significant advantage provided by the use of computers is 
that all parameters (e.g. modal logic, domain conditions, 
semantics) under which the announced results hold must be 
explicitly specified in the source code. This reduces the need 
for interpretation and the danger of misunderstandings. 
Current technology is increasingly ready to be embraced by 
those willing to practice computer-assisted theoretical 
philosophy \cite{oppenheimera11,rushby13}.

>>>>

Attempts to prove the
existence (or non-existence) of God by means of abstract, ontological
arguments are an old tradition in western philosophy, with contributions by several prominent philosophers, including St. Anselm of
Canterbury, Descartes and Leibniz. Kurt G{\"o}del and Dana Scott studied and improved this argument, bringing it to a mathematically more precise form, as a chain of axioms, lemmas and theorems in a second-order modal logic \cite{GoedelNotes,ScottNotes}, shown in Fig. \ref{fig:scott}.

@@

G\"{o}del defines God as a being who possesses all \emph{positive}
properties and states a few reasonable (but debatable) axioms that
such properties should satisfy.  The overall idea of G{\"o}del's proof
is in the tradition of Anselm's argument, who defined God as an
entity of which nothing greater can be conceived.  Anselm argued that
existence in the actual world would make such an assumed being even
greater (more perfect), hence, by definition, God must exist. However,
for Anselm existence was treated as a predicate and the possibility of
God's existence was assumed as granted.  These issues were criticized
by Kant and Leibniz, respectively, and they were addressed in the work
of G\"odel.

Nevertheless, G{\"o}del's work still leaves room for criticism.  In
particular, his axioms are so strong that, when assuming unrestricted
comprehension principles\footnote{A possible direction to remedy modal
  collapse, as studied e.g. by Koons
  \cite{koons06:_sobel_goedel_ontol_proof} is to impose restrictions
  on the domain of properties.}, 
%\marginpar{added footnote}
they entail a \emph{modal collapse}
\cite{Sobel1987,sobel2004logic}: everything that is the case is so
necessarily.  There has been an impressive body of recent and ongoing
work
(cf.~\cite{sobel2004logic,Fitting,anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings,bjordal99,fuhrmann05:_exist_notwen,Hajek2002,Hajek2008,ContemporaryBibliography}
and the references therein) proposing solutions for the modal
collapse.  The goal of this article is to discuss the modal
collapse from the point of view of the modal square of opposition. 
Ontological arguments typically rely on an inversion of the normal direction of entailment
in the modal square of opposition for one particular proposition (i.e. God's existence), 
and the modal collapse shows that this inversion in fact occurs for all propositions, resulting in a total collapse of the modal square of opposition.

@@

The modal collapse is an interesting example of philosophical 
controversy and dispute, to which we can apply Leibniz's idea 
of a \emph{calculus ratiocinator} brought to reality in the 
form of contemporary automated theorem provers. 
A significant advantage provided by the use of computers is 
that all parameters (e.g. modal logic, domain conditions, 
semantics) under which the announced results hold must be 
explicitly specified in the source code. Consequently, 
the danger of misunderstandings is reduced. 
Current technology is increasingly ready to be embraced by 
those willing to practice computer-assisted theoretical 
philosophy \cite{oppenheimera11,rushby13}.
